Notes on conformal manifolds

### 2025-10-03 Draft_MLFlatDirections.tex
Updates of the ref in intro, some paragraphs in sec 3, placement of the figures.

\begin{abstract}
	We investigate the structure of conformal manifolds around AdS$_3 \times S^3$ which lift from continuous flat directions in the scalar potential of gauged supergravity resulting from six-dimensional $\mathcal{N}=(1,1)$.
	Our approach combines numerical exploration and symbolic inference. For the latter, we develop a symbolic regression algorithm based on Annealed Importance Sampling combined with Sequential Monte Carlo, well-suited to uncovering polynomial constraints in high-dimensional parameter spaces. 
	The algorithm reconstructs a set of polynomial relations that provides an explicit analytic parametrization of a new family of solutions. 
\end{abstract}

@@

The AdS/CFT correspondence stands as one of the most profound dualities in theoretical physics, establishing a remarkable equivalence between string theory solutions on anti-de Sitter (AdS) spacetimes and conformal field theories (CFTs) living on their boundaries~\cite{Maldacena:1997re}. 
This correspondence has revolutionized our understanding of both quantum gravity and strongly coupled field theories, providing unprecedented insights into the holographic nature of gravity. 
Within this holographic framework, supergravity theories in AdS backgrounds serve as the low-energy effective descriptions of string theory compactifications, making them natural laboratories for exploring the gravitational side of the duality. On the other side, CFT’s appear in multiple scenarios, providing nice descriptions of phase transitions in statistical physics, and from a fundamental perspective, they are of utmost importance in quantum field theories (QFTs), where they describe fixed points of Renormalisation Group flow. 
From the latter perspective, an interesting question is  whether CFTs are isolated fixed points or instead belong to a continuous families. 
In that case, the space of continuous deformations that takes from one CFT to another is called a conformal manifold. 
These deformations are parametrised by exactly marginal operators, or in other words, operators that preserve the conformal symmetry, \textit{i.e.} whose $\beta$ functions exactly vanish. \ce{Bastien: changed here.}

@@

In this work, we demonstrate the viability of the aforementioned machine learning approach by applying it to a five-scalar subsector of a 13 scalar consistent truncation of six-dimensional non-chiral $\mathcal{N} = (1,1)$ supergravity on ${\rm AdS}_3 \times S^3$, or to type IIB supergravity on ${\rm AdS}_3 \times S^3 \times T^4$. Our methodology combines gradient descent sampling of the conformal manifold with a symbolic regression technique. There exists a large body of literature on symbolic regression, using methods from genetic programming \cite{koza1994genetic,virgolin2021improving,randall2022bingo,burlacu2019parsimony}, to deep learning \cite{petersen2019deep,kamienny2022end}, generative models \cite{valipour2021symbolicgpt}, diffusion models \cite{bastiani2025diffusion}, and equation learning with the nodes being symbolic operations \cite{2018arXiv180607259S}. A state-of-the-art algorithm is AIFeynman \cite{Udrescu:2019mnk}. It uses neural networks to identify structures in the dataset (such as translational symmetries, multiplicative separability, compositionality\dots) to recursively define simpler problems on which they can fit the solutions with polynomials and basic functions.  However, this method is slow, usually limited to low-dimensional spaces, and, by construction, can only fit one expression at a time. Alternatively, we develop here a symbolic regression algorithm based on Annealing Importance Sampling (AIS) \cite{neal1998annealedimportancesampling}, since this approach is much better suited to the study of geometric loci defined by intersecting polynomials in high-dimensional spaces

The study of the five-parameter potential is separated into several parts. We first use a gradient descent to efficiently sample the underlying conformal manifold. Combined with numerical analyses, including a local principal component analysis, we can identify the existence of a three-dimensional continuous family. We then use an AIS technique combined with a sequential Monte Carlo approach~(SMC)~\cite{del2006sequential} in order to do symbolic regression. As we will indeed demonstrate in the paper, there exist polynomial constraints on the 5 parameters, viewed as embedding coordinates, that project them onto the 3 dimensional manifolds. We manage to identify 8 of those constraints on the data, not all independent, which once solved  provide an explicit three-dimensional family of solutions.

\paragraph{}
The paper is organized as follows. In sec.~\ref{sec:sugra}, we establish the supergravity setup, presenting the scalar potential in its full complexity and motivating the restriction to five fields. Sec.~\ref{Sec:AIS-SMC} details our annealed importance sampling approach to symbolic regression. Our main results, both numerical an analytical, are presented in sec.~\ref{sec:results}, and sec.~\ref{sec:sugrasol} gives some details on the new supergravity solutions discovered by the numerical analysis. We conclude with prospects for extending this approach to higher-dimensional cases and its broader implications for systematic studies of conformal manifolds in holographic theories.

@@

	\subsection{General idea}
	The numerical sampling of the space of vacua (described in sec.~\ref{sec:graddes}) results in clouds of points embedded in a higher-dimensional space. 
	The aim of symbolic regression is to uncover interpretable mathematical expressions that best describe the embedding of these loci of solutions. 
	In the present context, the form of the supergravity potential \eqref{eq:scalarpotential} determining the solutions implies that the loci can be defined through polynomials $p_{m}$ such that $p_{m}\big(\vec{x}^{(i)}\big) = 0$, with $i \in \{1, \dots, n_{\rm points}\}$. Such polynomials will be called ``annihilating polynomials'' in the following.
	Finding annihilating polynomials involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. 
	Markov Chain Monte Carlo (MCMC) techniques such as Metropolis-Hastings or Gibbs sampling, while widely used, can struggle with poor mixing and often get trapped in local optima, especially in high-dimensional or multimodal spaces. \ce{Try to find a ref.}

	To address these challenges, we formulate the search for annihilating polynomials as a determination of a probability density on the space of polynomials that stresses higher probabilities on polynomials $p$ that minimise the error \ce{Bastien: changed here.}
	\begin{equation}
		\sum_{i=1}^{n_{\rm points}}p\big(\vec{x}^{(i)}\big)^{2}.
	\end{equation}
	Polynomials sampled from this density would thus be good approximations to annihilating polynomials. Such a probability density is \textit{a priori} hard to find, but can be approximated using algorithms based on Sequential Important Sampling~\cite{Doucet2001,Liu2004} such as Sequential Monte Carlo samplers~\cite{del2006sequential}. They are inspired from Important Sampling (IS)~\cite{osti_4441935,RobertCasella2004}, a method to parametrise expectation values with respect to a target distribution $d$, that can be evaluated pointwise, but from which one does not know how to draw samples. IS uses another, usually simpler, distribution $\pi$ (called importance distribution) that one knows how to draw from. It is based on the observation that the expectation value \ce{Change $\pi$ to $\eta$, $d$ to $\pi$.}
	\begin{equation}
		\mathbb{E}_{d}[f] = \int f(z)d(z)\ \d z
	\end{equation}
	for some function $f$ with respect to $d$ can be computed from the expectation $\mathbb{E}_{\pi}[w f]$, now computed with respect to the distribution $\pi$, with unnormalised weights $w = d/\pi$. Therefore, one can estimate $\mathbb{E}_{d}[f]$ from a Monte Carlo method: we draw a sample $\{z^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ from $\pi$ and use the weighted average
	\begin{equation}
		\mathbb{E}_{d}[f] \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} w\big(z^{(k)}\big)\,f\big(z^{(k)}\big)/\displaystyle \sum_{i=1}^{n_{\rm particles}} w\big(z^{(k)}\big).
	\end{equation}
	Stated differently, the target distribution $d$ can be approximated from a weighted collection of particles $\{z^{(k)},w^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ drawn from the importance distribution $\pi$ :
	\begin{equation}
		d(z) \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}\,\delta\big(z-z^{(k)}\big)/\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)},
	\end{equation}
	where $w^{(k)} = w\big(z^{(k)}\big)$. The sample admits $d$ as marginal distribution asymptotically when $n_{\rm particles}\rightarrow+\infty$. The accuracy of this estimation depends on the size of the sample and, crucially, on the choice of the distribution $\pi$, that needs to be close enough to $d$.

	However, providing a good importance distribution $\pi$ remains a hard problem. In order to fix this point, we employ a Sequential Monte Carlo sampler (SMC)~\cite{del2006sequential,DoucetTutorial,naesseth2024elementssequentialmontecarlo} enhanced by Annealed Importance Sampling (AIS)~\cite{neal1998annealedimportancesampling}. SMC aims at estimating $\pi$ sequentially from multiple intermediate distributions $\pi_{n}$ that smoothly transition from an initial, tractable distribution $\pi_{0}$ (chosen to be easy to sample from) to the complex distribution $\pi$. These distributions are approximated by weighted samples of particles $\big\{z^{(k)}_{n},w^{(k)}_{n}\big\}$ approximating $\pi_{n}$:
	\begin{equation} \label{eq:pifromweights}
		\pi_{n}(z) \simeq \sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}_{n}\,\delta(z-z^{(k)}_{n}),
	\end{equation}
	with  $\displaystyle \tilde{w}^{(k)}_n  = w^{(k)}_n/\sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}$ the normalised weights. The particles are incrementally deformed and reweighed using a Markov kernel to gradually sample the target distribution $d$. This process is guided by a temperature-like parameter that gradually emphasises the data likelihood, allowing for more efficient and tunable exploration of the polynomials landscape. The approximation procedure can be summed up as follows. We define the schedule of intermediate distributions $\pi_{n}$ using a temperature-like parameter, starting from an initial, easy to sample distribution $\pi_0$ which is parametrised by a set of particles and weights $\big\{z_0^{(k)},w_0^{(k)}\big\}$. We gradually mutate them with an annealing procedure to get a set of $\{z_n^{(k)},w_n^{(k)}\}$, which are used to parametrise $\pi_n$. As $n$ increases, the procedure converges such that the particles sample the distribution $\pi$, which itself is used to parametrise the target distribution $d$.

	This procedure has the disadvantage that the weights variance tends to increase, leading to weight degeneracy. AIS-SMC includes a resampling procedure, that focusses computational effort on high-probability regions.  This combination of importance sampling, mutation and resampling maintains diversity among the particles and prevents premature convergence to suboptimal models. These features make AIS-SMC particularly well-suited for symbolic regression tasks, where the search space is not only high-dimensional but also structured and discontinuous.

	\subsection{Detailed procedure}	\label{sec:detailedAISSMC}
	Let us now explain in more detail the procedure, based on ref.~\cite{del2006sequential,DoucetTutorial,naesseth2024elementssequentialmontecarlo}. 
	The goal is to obtain particles sampled from a probability distribution $d(z)$, where $z$ runs over a space of polynomials \ce{Here we should say what this space is, and stress that this space is finite dimensional.}, that assigns high probabilities to those polynomials that annihilate the data, and therefore will be used to identify their symbolic expressions. We reconstruct this density function $d$ by series of density functions $\pi_n(z)$, $n\in\llbracket0,n_{\rm epochs}\rrbracket$, such that $\pi_n\to d$ as $n\to n_{\rm epochs}$. On every annealing step, $\pi_n$ is defined in terms of an unnormalised density $\gamma_n$ and a normalisation constant as $\pi_n(z) = \gamma_n(z)/Z_n$. The unnormalised density at level $n$ is given in terms of a prior distribution $d_0(z)$ over the space of polynomials and a loss function $L(z)$, as
		\begin{equation}\label{eq:gamman}
			\gamma_n(z) = d_0(z) \: \exp \Big( -\beta_n L(z)\Big),
		\end{equation}
	where the inverse temperature constants $\beta_n$ are taken to evolve monotonically as $\beta_0 < \beta_1 <\dots< \beta_{n_{\rm epochs}}$. The importance given to each polynomial $z$ is thus updated at each step, with polynomials $z$ with high loss being disfavoured. This update is controlled by the inverse temperature constants, enabling deeper exploration during the early stages and protecting the procedure from being trapped in local minima. The loss function is chosen to be minimal on annihilating polynomials and its design will be discussed at the end of the section. We choose the prior distribution $d_{0}$ to be flat on the space of polynomials, giving equal importance to all of them, which makes easy the initial sampling of the weighted particles $\big\{z^{(k)}_{0},w^{(k)}_{0} = 1/n_{\rm particles}\big\}$. \ce{Change sentence.}

	At each epoch $n\geq1$, we do not directly approximate $\pi_{n}$, but rather a distribution $\tilde{\pi}_{n}\big(z_{0:n}\big) = \tilde{\gamma}_{n}\big(z_{0:n}\big)/\tilde{Z}_{n}$, with $z_{0:n} = \big(z_{0},\ldots,z_{n}\big)$, that describes the global trajectory of the particles until step~$n$. $\tilde{\gamma}_{n}$ is defined thanks to a backward propagation kernel $K\big(z_{i-1}\vert z_{i}\big)$:\footnote{Note that $\pi_{n}$ is a function on the cartesian product blabla.} \ce{Footnote.}
	\begin{equation}
		\tilde{\gamma}_{n}\big(z_{0:n}\big) = \gamma_{n}\big(z_{n}\big)\,\prod_{i=1}^{n} K\big(z_{i-1}\vert z_{i}\big).
	\end{equation}
	This way, the distribution $\tilde{\pi}_{n}$ describes $\pi_{n}$ as well as the path that followed the particles from epochs $0$ to $n$. This allows for more accuracy, correcting potential bias introduced by the mutations, and results in a setup that is more tunable.

	At the beginning of epoch $n\geq1$, $\tilde{\pi}_{n-1}$ is approximated \textit{via} Importance Sampling from an importance distribution $\eta_{n-1}$, \textit{i.e.} from a set of particles and weights $\big\{z_{n-1}^{(k)},\ w_{n-1}^{(k)}\big\}$ using eq.~\eqref{eq:pifromweights}, with $\big\{z_{n-1}^{(k)}\big\}$ drawn from $\eta_{n-1}$ and $w_{n-1}^{(k)} = \tilde{\gamma}_{n-1}\big(z_{0:n-1}^{(k)}\big)/\eta_{n-1}\big(z_{0:n-1}^{(k)}\big)$.\footnote{We define $\eta_{0} = \pi_{0} = \tilde{\pi}_{0} = d_{0}$.} \ce{Footnote.} The importance density $\eta_{n-1}$ gets mutated using a forward propagation Markov kernel $q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)$, defining the probability to get $z^{(k)}_{n}$ when the current state is $z^{(k)}_{n-1}$:
	\begin{equation}
		\eta_{n}\big(z_{0:n}\big) = \eta_{n-1}\big(z_{0:n-1}\big)\,q\big(z_{n}\vert z_{n-1}\big).
	\end{equation}
	We then use Importance Sampling to approximate $\tilde{\pi}_{n}$: we draw particles $\big\{z^{(k)}_{n}\big\}$ from $\eta_{n}$, and compute the weights
	% \begin{equation}
	% 	w_{n}^{(k)} = \frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\eta_{n}\big(z_{0:n}^{(k)}\big)} = \frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\eta_{n-1}\big(z_{0:n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)} = w_{n-1}^{(k)}\,\frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\tilde{\gamma}_{n-1}\big(z_{0:n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)}= w_{n-1}^{(k)}\,\frac{\gamma_{n}\big(z_{n}^{(k)}\big)\,L\big(z^{(k)}_{n-1}\vert z^{(k)}_{n}\big)}{\gamma_{n-1}\big(z_{n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)}.
	% \end{equation}
	\begin{equation} \label{eq:weightsupdate}
		w_{n}^{(k)} = \frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\eta_{n}\big(z_{0:n}^{(k)}\big)} = w_{n-1}^{(k)}\,\alpha_{n}^{(k)},
	\end{equation}
	with the incremental importance weights $\alpha_{n}^{(k)}$ defined as
	\begin{equation}	\label{eq: incremental_importance}
			\alpha^{(k)}_{n} =  \frac{\gamma_{n}\big(z_{n}^{(k)}\big)\,K\big(z^{(k)}_{n-1}\vert z^{(k)}_{n}\big)}{\gamma_{n-1}\big(z_{n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)} = \frac{\exp\Big(-\beta_{n+1}L\big(z_{n+1}^{(k)}\big)\Big)}{\exp\Big(-\beta_{n}L\big(z_{n}^{(k)}\big)\Big)}\,\frac{K\big(z^{(k)}_{n}\vert z^{(k)}_{n+1}\big)}{q\big(z^{(k)}_{n+1}\vert z^{(k)}_{n}\big)}.
	\end{equation}

>>>>

\begin{abstract}
	We investigate the structure of conformal manifolds around AdS$_3 \times S^3$ which lift from continuous flat directions in the scalar potential of gauged supergravity resulting from six-dimensional $\mathcal{N}=(1,1)$.
	Our approach combines numerical exploration and symbolic inference. For the latter, we develop a symbolic regression algorithm based on Annealed Sequential Monte Carlo samples, a combination of Annealed Importance Sampling and Sequential Monte Carlo samplers, well-suited to uncovering polynomial constraints in high-dimensional parameter spaces. 
	The algorithm reconstructs a set of polynomial relations that provides an explicit analytic parametrization of a new family of solutions. 
\end{abstract}

@@

The AdS/CFT correspondence stands as one of the most profound dualities in theoretical physics, establishing a remarkable equivalence between string theory solutions on anti-de Sitter (AdS) spacetimes and conformal field theories (CFTs) living on their boundaries~\cite{Maldacena:1997re}. 
This correspondence has revolutionized our understanding of both quantum gravity and strongly coupled field theories, providing unprecedented insights into the holographic nature of gravity. 
Within this holographic framework, supergravity theories in AdS backgrounds serve as the low-energy effective descriptions of string theory compactifications, making them natural laboratories for exploring the gravitational side of the duality. On the other side, CFT’s appear in multiple scenarios, providing nice descriptions of phase transitions in statistical physics, and from a fundamental perspective, they are of utmost importance in quantum field theories (QFTs), where they describe fixed points of Renormalisation Group flow. 
From the latter perspective, an interesting question is  whether CFTs are isolated fixed points or instead belong to a continuous families. 
In that case, the space of continuous deformations that takes from one CFT to another is called a conformal manifold. 
These deformations are parametrised by exactly marginal operators, or in other words, operators that preserve the conformal symmetry, \textit{i.e.} whose $\beta$ functions exactly vanish.

@@

In this work, we demonstrate the viability of the aforementioned machine learning approach by applying it to a five-scalar subsector of a 13 scalar consistent truncation of six-dimensional non-chiral $\mathcal{N} = (1,1)$ supergravity on ${\rm AdS}_3 \times S^3$, or to type IIB supergravity on ${\rm AdS}_3 \times S^3 \times T^4$. Our methodology combines gradient descent sampling of the conformal manifold with a symbolic regression technique. There exists a large body of literature on symbolic regression, using methods from genetic programming \cite{koza1994genetic,virgolin2021improving,randall2022bingo,burlacu2019parsimony}, to deep learning \cite{petersen2019deep,kamienny2022end}, generative models \cite{valipour2021symbolicgpt}, diffusion models \cite{bastiani2025diffusion}, and equation learning with the nodes being symbolic operations \cite{2018arXiv180607259S}. A state-of-the-art algorithm is AIFeynman \cite{Udrescu:2019mnk}. It uses neural networks to identify structures in the dataset (such as translational symmetries, multiplicative separability, compositionality\dots) to recursively define simpler problems on which they can fit the solutions with polynomials and basic functions.  However, this method is slow, usually limited to low-dimensional spaces, and, by construction, can only fit one expression at a time. Alternatively, we develop here a symbolic regression algorithm based on Annealed Sequential Monte Carlo sampler~\cite{zimmermann2021nested,syed2024optimisedannealedsequentialmonte} (combining the Annealed Importance Sampling~\cite{neal1998annealedimportancesampling} approach with a Sequential Monte Carlo sampler (SMC)~\cite{del2006sequential}), since this approach is much better suited to the study of geometric loci defined by intersecting polynomials in high-dimensional spaces.

The study of the five-parameter potential is separated into several parts. We first use a gradient descent to efficiently sample the underlying conformal manifold. Combined with numerical analyses, including a local principal component analysis, we can identify the existence of a three-dimensional continuous family. We then use an ASMC technique in order to do symbolic regression. As we will indeed demonstrate in the paper, there exist polynomial constraints on the 5 parameters, viewed as embedding coordinates, that project them onto the 3 dimensional manifolds. We manage to identify 8 of those constraints on the data, not all independent, which once solved  provide an explicit three-dimensional family of solutions.

\paragraph{}
The paper is organized as follows. In sec.~\ref{sec:sugra}, we establish the supergravity setup, presenting the scalar potential in its full complexity and motivating the restriction to five fields. Sec.~\ref{Sec:AIS-SMC} details our Annealed Sequential Monte Carlo sampler approach to symbolic regression. Our main results, both numerical an analytical, are presented in sec.~\ref{sec:results}, and sec.~\ref{sec:sugrasol} gives some details on the new supergravity solutions discovered by the numerical analysis. We conclude with prospects for extending this approach to higher-dimensional cases and its broader implications for systematic studies of conformal manifolds in holographic theories.

@@

	\subsection{General idea}
	The numerical sampling of the space of vacua (described in sec.~\ref{sec:graddes}) results in clouds of points embedded in a higher-dimensional space. 
	The aim of symbolic regression is to uncover interpretable mathematical expressions that best describe the embedding of these loci of solutions. 
	In the present context, the form of the supergravity potential \eqref{eq:scalarpotential} determining the solutions implies that the loci can be defined through polynomials $p_{m}$ such that $p_{m}\big(\vec{x}^{(i)}\big) = 0$, with $i \in \{1, \dots, n_{\rm points}\}$. Such polynomials will be called ``annihilating polynomials'' in the following.
	Finding annihilating polynomials involves navigating a vast, discrete, and often rugged search space of possible symbolic models, which poses significant challenges for traditional sampling methods. 
	Markov Chain Monte Carlo (MCMC) techniques such as Metropolis-Hastings or Gibbs sampling, while widely used, can struggle with poor mixing and often get trapped in local optima, especially in high-dimensional or multimodal spaces.

	To address these challenges, we formulate the search for annihilating polynomials as a determination of a probability density on the space of polynomials that stresses higher probabilities on polynomials $p$ that minimise the error
	\begin{equation}
		\sum_{i=1}^{n_{\rm points}}p\big(\vec{x}^{(i)}\big)^{2}.
	\end{equation}
	Polynomials sampled from this density would thus be good approximations to annihilating polynomials. Such a probability density is \textit{a priori} hard to find, but can be approximated using algorithms based on Sequential Important Sampling~\cite{Doucet2001,Liu2004} such as Sequential Monte Carlo samplers~\cite{del2006sequential}. They are inspired from Important Sampling (IS)~\cite{osti_4441935,RobertCasella2004}, a method to parametrise expectation values with respect to a target distribution $\pi$, that can be evaluated pointwise, but from which one does not know how to draw samples. IS uses another, typically simpler, distribution $\eta$ (called importance distribution) that one knows how to draw from. It is based on the observation that the expectation value
	\begin{equation}
		\mathbb{E}_{\pi}[f] = \int f(z)\pi(z)\ \d z
	\end{equation}
	for some function $f$ with respect to $\pi$ can be computed from the expectation $\mathbb{E}_{\eta}[w f]$, now computed with respect to the distribution $\eta$, with unnormalised weights $w = \pi/\eta$. Therefore, one can estimate $\mathbb{E}_{\pi}[f]$ from a Monte Carlo method: we draw a sample $\{z^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ from $\eta$ and use the weighted average
	\begin{equation}
		\mathbb{E}_{\pi}[f] \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} w\big(z^{(k)}\big)\,f\big(z^{(k)}\big)\Bigg/\displaystyle \sum_{i=1}^{n_{\rm particles}} w\big(z^{(k)}\big).
	\end{equation}
	Stated differently, the target distribution $\pi$ can be approximated from a weighted collection of particles $\{z^{(k)},w^{(k)}\}_{k\in\llbracket 1, n_{\rm particles}\rrbracket}$ drawn from the importance distribution $\eta$ :
	\begin{equation}
		\pi(z) \simeq \displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)}\,\delta\big(z-z^{(k)}\big)\Bigg/\displaystyle \sum_{k=1}^{n_{\rm particles}} w^{(k)},
	\end{equation}
	where $w^{(k)} = w\big(z^{(k)}\big)$. The sample admits $\pi$ as marginal distribution asymptotically when $n_{\rm particles}\rightarrow+\infty$. The accuracy of this estimation depends on the size of the sample and, crucially, on the choice of the importance distribution $\eta$, that needs to be close enough to $\pi$.

	However, providing a good importance distribution $\eta$ remains a hard problem. In order to fix this point, we employ an Annealed Sequential Monte Carlo sampler (ASMC)~\cite{zimmermann2021nested,syed2024optimisedannealedsequentialmonte}, \textit{i.e.} a Sequential Monte Carlo sampler (SMC)~\cite{del2006sequential,DoucetTutorial,naesseth2024elementssequentialmontecarlo} enhanced by Annealed Importance Sampling (AIS)~\cite{neal1998annealedimportancesampling}. SMC aims at estimating $\pi$ sequentially from multiple intermediate distributions $\pi_{n}$ that smoothly transition from an initial, tractable distribution $\pi_{0}$ (chosen to be easy to sample from) to the complex distribution $\pi$. These distributions are approximated by weighted samples of particles $\big\{z^{(k)}_{n},w^{(k)}_{n}\big\}$ approximating $\pi_{n}$:
	\begin{equation} \label{eq:pifromweights}
		\pi_{n}(z) \simeq \sum_{k=1}^{n_{\rm particles}} \tilde{w}^{(k)}_{n}\,\delta(z-z^{(k)}_{n}),
	\end{equation}
	with  $\displaystyle \tilde{w}^{(k)}_n  = w^{(k)}_n/\sum_{k=1}^{n_{\rm particles}} w^{(k)}_{n}$ the normalised weights. The particles are incrementally deformed and reweighed using a Markov kernel to gradually sample the target distribution $\pi$. This process is guided by a temperature-like parameter that gradually emphasises the data likelihood, allowing for more efficient and tunable exploration of the polynomials landscape. The approximation procedure can be summed up as follows. We define the schedule of intermediate distributions $\pi_{n}$ using a temperature-like parameter, starting from an initial, easy to sample distribution $\pi_0$ which is parametrised by a set of particles and weights $\big\{z_0^{(k)},w_0^{(k)}\big\}$. We gradually mutate them with an annealing procedure to get a set of $\{z_n^{(k)},w_n^{(k)}\}$, which are used to parametrise $\pi_n$. As $n$ increases, the procedure converges such that the particles sample the target distribution $\pi$.

	This procedure has the disadvantage that the weights variance tends to increase, leading to weight degeneracy. ASMC includes a resampling procedure, that focusses computational effort on high-probability regions.  This combination of importance sampling, mutation and resampling maintains diversity among the particles and prevents premature convergence to suboptimal models. These features make ASMC particularly well-suited for symbolic regression tasks, where the search space is not only high-dimensional but also structured and discontinuous.

	\subsection{Detailed procedure}	\label{sec:detailedAISSMC}
	Let us now explain in more detail the procedure, based on ref.~\cite{del2006sequential,DoucetTutorial,naesseth2024elementssequentialmontecarlo,Dai03072022}. 
	The goal is to obtain particles sampled from a probability distribution $\pi(z)$, where $z$ runs over a space of polynomials $E$, that assigns high probabilities to those polynomials that annihilate the data, and therefore will be used to identify their symbolic expressions. We restrict ourselves to polynomials of maximum degree ${\rm max}_{\rm degree}$. Given the number of variables $n_{\rm var}$, this fixes the number of different possible monomials to $\binom{{\rm max}_{\rm degree}+n_{\rm var}}{{\rm max}_{\rm degree}}$. We further restrict $E$ to include only polynomials with a maximum number of ${\rm max}_{\rm mon}$ monomials. The space $E$ is then finite dimensional.

	We reconstruct the density function $\pi$ by series of density functions $\pi_n(z)$, $n\in\llbracket0,n_{\rm epochs}\rrbracket$, such that $\pi_n\to \pi$ as $n\to n_{\rm epochs}$. On every annealing step, $\pi_n$ is defined in terms of an unnormalised density $\gamma_n$ and a normalisation constant as $\pi_n(z) = \gamma_n(z)/Z_n$. The unnormalised density at level $n$ is given in terms of a prior distribution $\pi_0(z)$ over the space $E$ of polynomials and a loss function $L(z)$, as
		\begin{equation}\label{eq:gamman}
			\gamma_n(z) = \pi_0(z) \: \exp \Big( -\beta_n L(z)\Big),
		\end{equation}
	where the inverse temperature constants $\beta_n$ are taken to evolve monotonically as $\beta_0 = 0 < \beta_1 <\dots< \beta_{n_{\rm epochs}}$. The importance given to each polynomial $z$ is thus updated at each step, with polynomials $z$ with high loss being disfavoured. This update is controlled by the inverse temperature constants, enabling deeper exploration during the early stages and protecting the procedure from being trapped in local minima. The loss function is chosen to be minimal on annihilating polynomials and its design will be discussed at the end of the section. We choose the prior distribution $\pi_{0}$ to be flat on the space of polynomials $E$, giving equal importance to all of them, which makes easy the initial sampling of the weighted particles $\big\{z^{(k)}_{0},w^{(k)}_{0} = 1/n_{\rm particles}\big\}$.\footnote{Note that with this choice there is a higher probability to draw polynomials with multiple monomials.} The random draw is done on the monomials of each polynomials, and the coefficients are chosen randomly in the range $[-2,2]$.

	At each epoch $n\geq1$, we do not directly approximate $\pi_{n}$, but rather a distribution $\tilde{\pi}_{n}\big(z_{0:n}\big) = \tilde{\gamma}_{n}\big(z_{0:n}\big)/\tilde{Z}_{n}$ defined on $E^{n+1}$, with $z_{0:n} = \big(z_{0},\ldots,z_{n}\big)$, that describes the global trajectory of the particles until step $n$. $\tilde{\gamma}_{n}$ is defined thanks to a backward propagation kernel $K\big(z_{i-1}\vert z_{i}\big)$:
	\begin{equation}
		\tilde{\gamma}_{n}\big(z_{0:n}\big) = \gamma_{n}\big(z_{n}\big)\,\prod_{i=1}^{n} K\big(z_{i-1}\vert z_{i}\big).
	\end{equation}
	This way, the distribution $\tilde{\pi}_{n}$ describes $\pi_{n}$ as well as the path that followed the particles from epochs $0$ to $n$. $\tilde{\pi}_{n}$ evaluates not only the probability that the particle is in a given state $z_{n}$, but also the probability that the particle has followed the path from $z_{0}$ to $z_{n}$. This allows for more accuracy, correcting potential bias introduced by the mutations, and results in a setup that is more tunable.

	At the beginning of epoch $n\geq1$, $\tilde{\pi}_{n-1}$ is approximated \textit{via} Importance Sampling from an importance distribution $\eta_{n-1}$, \textit{i.e.} from a set of particles and weights $\big\{z_{n-1}^{(k)},\ w_{n-1}^{(k)}\big\}$ using eq.~\eqref{eq:pifromweights}, with $\big\{z_{n-1}^{(k)}\big\}$ drawn from $\eta_{n-1}$ and $w_{n-1}^{(k)} = \tilde{\gamma}_{n-1}\big(z_{0:n-1}^{(k)}\big)/\eta_{n-1}\big(z_{0:n-1}^{(k)}\big)$.\footnote{We define $\eta_{0} = \pi_{0} = \tilde{\pi}_{0}$.} The importance density $\eta_{n-1}$ gets mutated using a forward propagation Markov kernel $q\big(z_{n}\vert z_{n-1}\big)$, defining the probability to get $z_{n}$ when the current state is $z_{n-1}$:
	\begin{equation}
		\eta_{n}\big(z_{0:n}\big) = \eta_{n-1}\big(z_{0:n-1}\big)\,q\big(z_{n}\vert z_{n-1}\big).
	\end{equation}
	We then use Importance Sampling to approximate $\tilde{\pi}_{n}$: we draw particles $\big\{z^{(k)}_{n}\big\}$ from $\eta_{n}$, and compute the weights
	% \begin{equation}
	% 	w_{n}^{(k)} = \frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\eta_{n}\big(z_{0:n}^{(k)}\big)} = \frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\eta_{n-1}\big(z_{0:n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)} = w_{n-1}^{(k)}\,\frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\tilde{\gamma}_{n-1}\big(z_{0:n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)}= w_{n-1}^{(k)}\,\frac{\gamma_{n}\big(z_{n}^{(k)}\big)\,L\big(z^{(k)}_{n-1}\vert z^{(k)}_{n}\big)}{\gamma_{n-1}\big(z_{n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)}.
	% \end{equation}
	\begin{equation} \label{eq:weightsupdate}
		w_{n}^{(k)} = \frac{\tilde{\gamma}_{n}\big(z_{0:n}^{(k)}\big)}{\eta_{n}\big(z_{0:n}^{(k)}\big)} = w_{n-1}^{(k)}\,\alpha_{n}^{(k)},
	\end{equation}
	with the incremental importance weights $\alpha_{n}^{(k)}$ defined as
	\begin{equation}	\label{eq: incremental_importance}
			\alpha^{(k)}_{n} =  \frac{\gamma_{n}\big(z_{n}^{(k)}\big)\,K\big(z^{(k)}_{n-1}\vert z^{(k)}_{n}\big)}{\gamma_{n-1}\big(z_{n-1}^{(k)}\big)\,q\big(z^{(k)}_{n}\vert z^{(k)}_{n-1}\big)} = \frac{\exp\Big(-\beta_{n+1}L\big(z_{n+1}^{(k)}\big)\Big)}{\exp\Big(-\beta_{n}L\big(z_{n}^{(k)}\big)\Big)}\,\frac{K\big(z^{(k)}_{n}\vert z^{(k)}_{n+1}\big)}{q\big(z^{(k)}_{n+1}\vert z^{(k)}_{n}\big)}.
	\end{equation}
